---
title: "DMC_alltest Visualization"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#package setup

```{r}

#install.packages('ggforce')
library(RColorBrewer)
library(kableExtra)
library(readr)
library(tidyverse)
library(reshape2)
library(cowplot)
library(ggpubr)
library(egg)
library(ggplot2)
library(cowplot)
library(ggforce)
library(scales)
```

# Load data

```{r}
# parent directory
base_path <- file.path("Z:", "Victor", "Projects", "NIH - cells", "data", "main", "interim")

# sample data
pDat <- readRDS(file.path(base_path, '3_1_pDat_filt.rds'))

# methylation data
betas <- readRDS(file.path(base_path, '3_1_betas_filt.rds'))

# annotation
anno <- readRDS(file.path('Z:', 'Victor', 'Repositories', 'EPIC_annotation',
                          'hg19_epic_annotation.rds'))

#color code
color_code <- readRDS(file.path(base_path, '2_3_color_code.rds'))
color_code_tissue <- setNames(color_code$Colors_Tissue, color_code$label)
colors <- color_code_tissue[unique(pDat$Tissue)]

#list of genes for search list
gene <- tibble(gene = str_split(anno$genes_symbol, ', ') %>%
                 unlist() %>%
                 unique() %>%
                 na.omit() %>%
                 sort())

#pull out cpg name for search list use
cpg <- unique(anno$cpg) 
cpg <- cpg[cpg %in% rownames(betas)] %>% as_tibble()
```

# 1. datatable which shows cpg annotation

## Main function

```{r}
#datatable which shows cpg annotation (when inputs have both cpgs and genes)
cpg_info_total <- function(cpg_name = NULL, gene_name = NULL){
    if (is.null(cpg_name)){
      info <- cpg_info_gene(gene_name)
    }
    else if (gene_name == ''){
      info <- cpg_info(cpg_name)
    }
    else {
      stop('please enter either cpgs or a gene')
    }
    
  info
}
```

##when input only cpgs
```{r}
#datatable which shows cpg annotation
anno1 <- anno %>% 
  select(cpg, chr, start, end, 
         cpg_id, enhancers_id, genes_id, genes_symbol, genes_tx_id, pmd_id, 
         imprinted_dmr_general, imprinted_dmr_placenta)

cpg_info <- function(cpg_name){

  cpg <- filter(anno1, cpg %in% cpg_name) %>% arrange(desc(start))

cpg
}
```

##when input only genes
```{r}
#datatable shows all cpg sites related to input gene
cpg_info_gene <- function(gene_name){

  #get cpgs sites related to input genes
  cpg_name <- fn_cpg(gene_name)
  
  cpg <- filter(anno1, cpg %in% cpg_name)%>% arrange(desc(start))
  cpg
}
```


# 2. Main gene/cpg plot with tracks

This is the middle plot with all the annotation tracks.

```{r}
stat_plot <- function(cpg_name = NULL, gene_name = NULL){
  
  ######## FILTER ANNOTATION TO RELEVANT CPGS, DEPENDING ON INPUT #################
  if (is.null(cpg_name)){
     
    # If a gene symbol is supplied,
    # filter to gene in annotation
    anno_gene <- anno %>%
      filter(grepl(paste0('\\<', gene_name, '\\>'), genes_symbol),
             cpg %in% rownames(betas)) %>%
      arrange(desc(start))
    
    # stop if gene symbol is invalid
    if (nrow(anno_gene) == 0) {
      stop('No cpgs found, maybe they were filtered out.')
    }
  } else if (gene_name == ''){
      
      # If a cpg is provided,
      # filter annotation to cpgs
      anno_gene <- anno %>%
        filter(cpg %in% cpg_name,
               cpg %in% rownames(betas)) %>% 
        arrange(chr, desc(start))
      
      # stop if no cpgs were found
      if (nrow(anno_gene) == 0) {
        stop('None of the selected CpGs exist in our processed data.')
      }
      
      if (length(cpg_name) < length(intersect(cpg_name, rownames(betas)))) {
        print(paste0('The following selected CpGs do not exist in our processed data:\n', 
                     setdiff(cpg_name, rownames(betas))))
      }
      
  } else {
      stop('Enter valid cpg(s) OR one specific gene.')
  }
  
  ######## SETUP INPUT TO PLOTS ###############
  #
  # now we wrangle the betas/annotations/pdata information into a format usable for plotting
  #
  betas_gene <- t(betas[anno_gene$cpg,,drop = FALSE]*100) %>% as_tibble %>% 
    mutate(Sample_Name = colnames(betas)) %>%
    
    # reshape into longer format
    pivot_longer(cols = -Sample_Name,
                 names_to = 'cpg', 
                 values_to = 'beta')  %>%
    
    # add tissue and trimester info
    left_join(pDat %>% select(Sample_Name, Trimester, Tissue), by = 'Sample_Name') %>%
    
    # calculate mean and sd for each cpg for each group
    group_by(Tissue, Trimester, cpg) %>%
    summarize(mean = mean(beta),
              sd = sd(beta)) %>%
    mutate(lower = mean-sd, upper = mean+sd) %>%
    
    # add cpg info
    left_join(anno_gene, by = 'cpg') %>%
    
    # order on cpg position
    ungroup() %>%
    arrange(desc(start)) %>%
    mutate(cpg = factor(cpg, levels = unique(cpg)),
           Trimester_Tissue = paste0(Trimester, ' - ', Tissue))
  
  ##### GENERATE INDIVIDUAL PLOT TRACKS #####
  font_size <- 12
  point_size <- 1
  
  # Generate  methylation track
  p1 <- betas_gene %>%
    ggplot() +
    
    # geom_ribbon is to generate the alternating shaded background
    geom_ribbon(aes(x = as.numeric(cpg), ymin = 0, ymax = 25),
                fill = 'grey', alpha = 0.25)+
    geom_ribbon(aes(x = as.numeric(cpg), ymin = 50, ymax = 75),
                fill = 'grey', alpha = 0.25)+
    
    # geom_linerange and geom_point is for the actual methylation data
    geom_linerange(alpha = 0.5, size = point_size, 
                   aes(x = cpg, ymin =lower, ymax = upper, color = Tissue),
                   show.legend = FALSE) +
    geom_point(alpha = 0.75, aes(x = cpg, y = mean, color = Tissue), size = point_size*2) +
    
    # we want to facet by trimester
    facet_wrap(vars(Trimester), ncol = 1, labeller = label_both) +
    
    # cosmetics
    theme_bw(base_size = font_size) +
    theme(axis.text.x = element_blank(), 
          axis.text.y = element_text(size = 12),
          axis.ticks.x = element_blank(),
          axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0, size = 16),
          
          legend.text = element_text(size = 14),
          
          panel.grid.minor.y = element_blank(),
          panel.grid.major.y = element_blank(),
          panel.spacing.y = unit(0.5, 'cm'),
          
          strip.background = element_blank(),
          strip.text = element_text(face = 'bold', hjust = 0),
          
          plot.margin=margin(l=-0.1, unit="cm")) +
    scale_y_continuous(limits = c(0, 100), expand = c(0, 0), labels = function(x)paste0(x, '%')) +
    scale_x_discrete(expand = c(0.01,0.01))+
    scale_color_manual(values= color_code_tissue[unique(betas_gene$Tissue)],
                       guide = guide_legend(override.aes = list(size = 4))) +
    labs(y = 'DNA\nmethylation', 
         x = '', 
         color = '', 
         title = if_else(!is.null(gene_name), gene_name, '')) 
    
  # plot cpg annotations
  p2 <- betas_gene %>% 
    
    # First we need to process the annotation data
    mutate(# absent/presence for different genomic elements
           enhancer = !is.na(enhancers_id),
           pmd = !is.na(pmd_id),
           imprinted_gene_placenta = !is.na(imprinted_gene_placenta),
           imprinted_gene_general = !is.na(imprinted_gene_general),
           imprinted_dmr_general = !is.na(imprinted_dmr_general),
           imprinted_dmr_placenta = !is.na(imprinted_dmr_placenta)) %>%
    mutate_if(is.logical, as.character) %>%
    select(cpg, enhancer, pmd, 
           imprinted_dmr_general, imprinted_dmr_placenta, cpg_id, start) %>%
    
    # reshape
    pivot_longer(cols = c(-cpg, -start),
               names_to = 'cpg_element', 
               values_to = 'presence') %>%
    distinct() %>%
    
    # arrange plot orde for each track
    arrange(desc(start)) %>%
    mutate(cpg = factor(cpg, levels = unique(cpg)),
           cpg_element = factor(cpg_element, 
                                levels = c('enhancer', 'pmd', 'imprinted_dmr_general',
                                           'imprinted_dmr_placenta', 'cpg_id')),
           presence = factor(ifelse(presence == "TRUE", 'Present',
                                    ifelse(presence == "FALSE", 'Absent', presence)),
                             levels = c('sea', 'shelf', 'shore', 'island',
                                        'Present', 'Absent', ' ', '  ')),
           # for facet label
           group = 'Annotations') %>%
    
    # plot code
    ggplot(aes(x = cpg, y = cpg_element, fill = presence)) +
    geom_tile(color = 'white') +
    guides(fill=guide_legend(ncol=2, override.aes = list(colour = 'white'))) +
    facet_wrap(vars(group), ncol = 2) +
    theme_bw(base_size = font_size) +
    theme(axis.text.x = element_blank(),
          axis.ticks = element_blank(),
          axis.text.y = element_text(vjust = 0.5),
          
          legend.text = element_text(size = 14),
          
          plot.margin = margin(l=-0.1,unit="cm"),
          
          strip.background = element_blank(),
          strip.text = element_text(face = 'bold', hjust = 0)) +
    scale_fill_manual(values = c('Present' = '#cccccc', 'Absent' = '#f7f7f7', 
                                 ' ' = 'white', '  ' = 'white',
                                 'sea' = '#ffffcc', 'shelf' = '#a1dab4', 'shore' = '#41b6c4',
                                 'island' = '#225ea8'), drop = F,
                      labels = stringr::str_to_title) +
    scale_y_discrete(expand = c(0,0),
                     labels = function(x)gsub('cpg_id', 'Relation to CpG Islands',
                       gsub('imprinted_dmr_placenta', 'Imprinted DMR (Placenta)',
                       gsub('imprinted_dmr_general', 'Imprinted DMR (General)',
                       gsub('pmd', 'Placental PMD',
                       gsub('enhancer', 'Enhancer', 
                            x)))))) +
    scale_x_discrete(expand = c(0,0)) +
    labs(x = '', y = '', fill = '')
  
  # transcript annotations
  
  # cpgs can have more than one mapping to a gene element per transcript
  # cpgs mapped to intronexonboundaries can also be part of exons and introns
  # 5 UTR and 3 UTR can also overlap with exons and introns
  
  # If there are multiple mappings then only the following will be displayed in order of priority:
  # 5UTR / 3UTR, exon / intron, intronexonboundary
  
  p3 <- betas_gene %>% 
    select(cpg, end, genes_id, genes_tx_id, genes_symbol) %>%
    distinct() %>%
    
    # split comma separate list and put each entry into a new row
    mutate(genes_tx_id = str_split(genes_tx_id, ', '),
           genes_id = str_split(genes_id, ', '),
           genes_symbol = str_split(genes_symbol, ', ')) %>%
    unnest(c(genes_tx_id, genes_id, genes_symbol)) %>%
    
    # paste together transcript ID and genes symbol
    mutate(genes_tx_id_symbol = paste0(genes_tx_id, ' (', genes_symbol, ')')) %>%
    
    # order transcripts by position of start
    arrange(end) %>%
    mutate(genes_tx_id_symbol = 
             factor(genes_tx_id_symbol, levels = unique(genes_tx_id_symbol)) %>%
             fct_explicit_na()) %>%
    
    # fill all combinations of cpg and gene_tx_id with NA if missing, so that they show in plot
    complete(cpg, genes_tx_id_symbol) %>%
    
    # filter out to unique mappings using the priority described above
    # strategy is to order by factor level and take the first occurence
    mutate(genes_id = factor(genes_id, 
                             levels = c('1to5kb', 'promoter', '5UTR', '3UTR', 'exon',
                                        'intron', 'intronexonboundary', 'intergenic'))) %>%
    
    group_by(cpg, genes_tx_id_symbol) %>%
    arrange(genes_id)  %>%
    dplyr::slice(1) %>%
    
    # reorder factor levels for appearance in plot legend
    mutate(genes_id = factor(genes_id, 
                             levels = c('1to5kb', 'promoter', '5UTR', 'exon', 'intron',
                                        'intronexonboundary', '3UTR', 'intergenic')),
           group = 'Transcripts') %>%
    
    #plot
    ggplot(aes(x = cpg, y = genes_tx_id_symbol, fill = genes_id)) +
    geom_tile(color = 'white', alpha = 0.6) +
    facet_wrap(vars(group), ncol = 1) +
    guides(fill=guide_legend(ncol=1, override.aes = list(alpha=0.7, color = 'white'))) +
    theme_bw(base_size = font_size) +
    theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1), 
          axis.text.y = element_text(vjust = 0.5),
          axis.ticks = element_blank(),
          
          plot.margin = margin(l=-0.1,unit="cm"),
          
          legend.text = element_text(size = 14),
          
          panel.grid = element_blank(),
          
          strip.background = element_blank(),
          strip.text = element_text(face = 'bold', hjust = 0)) +
    scale_fill_brewer(na.value = '#f7f7f7', breaks = c('1to5kb', 'promoter', '5UTR', 'exon', 
                                                        'intron', 'intronexonboundary', '3UTR',
                                                        'intergenic'),
                      palette = 'Paired', direction = -1) +
    scale_y_discrete(expand = c(0,0), 
                     labels = function(x)(if_else(x == '(Missing)', '', x))) +
    scale_x_discrete(expand = c(0,0)) +
    coord_cartesian(clip = "off") +
    labs(x = '', y = '', fill = '')
  
  # the size of each plot depends on the number of unique transcripts
  # More transcripts means a longer transcript track
  # so the plot height depends on the number of transcripts
  
  # number of transcripts
  n_trans <- betas_gene$genes_tx_id %>% str_split(', ') %>% unlist() %>% unique %>% length()
  
  # size of p1
  p1_h <- 35
  
  # combine tracks plot
  ggarrange(p1, p2, p3, ncol = 1, heights = c(p1_h-(n_trans-7), 7, n_trans))
  }

```

# 3. CpG boxplots

These functions define the individual CpG plots when the user selects specific CpGs.

## Data wrangling functions

These set of functions help extract the relevant information given user input, and then manipulates the data into a format usable by the plotting functions.

### Given a gene symbol, return associated cpg IDs

```{r}
fn_cpg <- function(gene_name){
  cpg_name <- anno %>%
    filter(cpg %in% rownames(betas),
           grepl(gene_name, genes_symbol)) %>%
    pull(cpg) %>% 
    unique()
  
  cpg_name
}
```

### Given cpgs, pull out DNA methylation from betas dataframe

Given a cpg, pull the associated DNA methylation at these cpgs, along with tissue and trimester information.

```{r}
sample_info_cpg <- function(cpg_name){
    
    cpg <- betas[cpg_name,]
    
    #when have only one cpg input
    if (length(cpg_name) ==1){
    cpg <- as.data.frame(cpg) %>% t()
    cpg <- cpg[,colnames(cpg) %in% pDat$Sample_Name]
    cpg <- as.data.frame(cpg)
    colnames(cpg) <- cpg_name
    cpg <- cpg %>% cbind(pDat) %>% gather(cpg, betas, contains('cg'))
    }
    
    #when have more than one cpg inputs
    else{
    cpg <- cpg[,colnames(cpg) %in% pDat$Sample_Name]
    cpg <- cpg %>% t() %>% cbind(pDat) %>% gather(cpg, betas, contains('cg'))
    }
    cpg <- cpg %>% 
      left_join(anno, by = 'cpg') %>% 
      arrange(desc(start)) %>% 
      mutate(cpg = factor(cpg, levels = unique(cpg)))
    cpg
}
```

Given a gene, pull the relevant cpgs, then betas + tissue/trimester information

```{r}
sample_info_gene <- function(gene_name){
  #get related cpg sites
  cpg_name <- fn_cpg(gene_name)
  #get sample cell types and tri
  cpg_info <- sample_info_cpg(cpg_name) 
  cpg_info
}
```

## Plot function

```{r}
#boxplots which show beta values based on cell types
boxplot_individual<- function(cpg_name){
  
    #given cpg, get sample betas, cell type  and trimester
    cpg <- sample_info_cpg(cpg_name)
    
    ggplot(cpg, aes(x=Trimester, y = betas, color = Tissue)) +
      geom_boxplot(outlier.shape = NA)+
      geom_jitter(alpha = 0.9, show.legend = F, 
                  position = position_jitterdodge(jitter.width = 0.05)) +
      facet_wrap_paginate(.~cpg)+ 
      theme_bw(base_size = 16)+
      theme(axis.text.x = element_text(angle = 45, vjust = 1, hjust =1),
            axis.title.y = element_text(angle = 0, vjust = 0.5, hjust = 0))+
      labs(x = ' ', y ='DNA\nmethylation', color = '') +
      scale_y_continuous(limits = c(0, 1), labels = scales::percent) +
      scale_fill_manual(values = colors) +
      scale_color_manual(values = colors) 
}
```
